---
title: "04_huva_gene_experiment"
output: html_document
---

### Load huva

```{r setup, warning=F}
library(huva)
library(huva.db)
library(useful)
```

# huva experiment

```{r}
gene_name <- "MYD88"
```

```{r, warning=F}
set.seed(1234)

exper <- run_huva_experiment(data = huva.db, 
                                      gene = gene_name, 
                                      quantiles = 0.10, 
                                      gs_list = hallmarks_V7.2,
                                      summ = T, 
                                      datasets_list = NULL, 
                                      adjust.method = "BH")
```

# Data Export

## Export the filtered expression table for the ImmVar CD4+ T cell dataset
```{r}
exper$ImmVar$data$ImmVar_CD4T
```

## Export the filtered annotation table for the CEDAR CD8+ T cell dataset
```{r}
exper$CEDAR$anno$CEDAR_CD8T
```

## Export DE genes table from the CEDAR monocytes dataset
```{r}
exper$CEDAR$DE_genes$CEDAR_CD14M
```

## Export the ranked gene list from the monocytes dataset
```{r}
exper$ImmVar$Rank_genelist$ImmVar_CD14M
```

## Export GSEA results from the CEDAR Granulocytes dataset
```{r}
exper$CEDAR$gsea$CEDAR_CD15G
```

## Export filtered cell count metadata table from the FG500 PBMC dataset
```{r}
exper$FG500$metadata$FG500_whole_blood_cellcount
```

## Export summary statistics from CEDAR CD8+ T cell sample annotation
```{r}
exper$summary$anno$CEDAR_CD8T
```

##  Export summary statistics from FG500 PBMC cytokines secretion metadata
```{r}
exper$summary$metadata$FG500_whole_blood_cytokines
```

# Session Info
```{r}
info <- sessionInfo()

info
```

